Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers

### 2025-01-18 endofquarterpaper.qmd
edits

The Upper Paleolithic is characterized by a rise in early Homo sapiens, and a decline for other hominids. During this time, we see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe. One of the most distinctive qualities of the Upper Paleolithic is the large rise in artwork available from that era, both parietal and otherwise. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012].

We still lack a complete knowledge of the cultural diversity and movement patterns of the early modern humans from this era, despite a great deal of research into the subject. In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian technocomplex. The basis of this analysis is the idea that variations on signs can serve as an indicator of group membership and cultural similarity. This approach is supported by both ethnographic evidence throughout history, as well as similar archaeological studies done on ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d]. In this study, we address the following questions: Are cultural groups evident in the distribution of geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?

<!--BM: our position is that cultural groups exist, our contribution is to see if/how they are evident in the sign data  -->

@@

Our understanding of the patterns of cultural and genetic diversity among the hominids of the Upper Paleolithic remains incomplete. Previous works have performed various cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity. The studies have primarily relied upon personal ornamentation and parietal art, with few taking mobile objects into account.

@@

Our use of geometric signs in this study is supported by @Sauvet_2018, in which used quadrilateral geometric parietal signs are used to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain. The authors identified and classified the signs based on variations in form, then performed a factorial correspondence analysis and ascending hierarchical clustering upon the data in order to identify two clusters, which they argued represents distinct and separate groups of hominids. In their study, @Sauvet_2018 made a case for the relevance of signs as identity markers, pointing to both the relevant qualities of signs, which are similar to those explained above, and the case studies throughout history where they served a similar purpose.

@@

abundance_graph <- ggplot(abundance_data) +
  aes(x = reorder(site_name, sign_total),
      y = sign_total) +
  geom_col() +
  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) +
  labs(x = "Site name",
       y = "Sign count")

library(cowplot)
plot_grid(date_graph, abundance_graph, artifact_map,
          labels = "AUTO")

@@

The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance. Assuming that this framework is true, we would hypothetically see patterns of cultural features of Upper Paleolithic hominids that are dispersed according to geography. Conversely, if geographic distance cannot account for the distribution of the sign types, we would have to assume that other factors are at play, namely, movement of groups or the creation of social networks.

@@

In this paper, we set out to ask if identifiable and relevant patterns existed in the data, and if so, what could they tell us about the patterns of diversity and movement among Upper Paleolithic hominids?

@@

However, this preliminary research has been able to produce some conclusions to be reasonably confident in. Firstly, given that we know definitively that distribution and variation of sign types is not the result of geographic distance, we can conclude that the data is likely relevant to understanding patterns of cultural diversity. Secondly, we have found clusters of sites that correspond in sign type, possibly indicating mobility of specific cultural groups. These conclusions contribute to, and broaden our current understanding of cultural and biological diversity among Upper Paleolithic hominids in Europe.

>>>>

One of the most distinctive qualities of the Upper Paleolithic is the large rise in available artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans. During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].

We still lack a complete knowledge of the cultural diversity and movement patterns of the early modern humans from this era, despite a great deal of research into the subject. In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian technocomplex. The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity. This approach is supported by both ethnographic evidence throughout history, as well as similar archaeological studies done on ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d]. In this study, we address the following questions: Are cultural groups evident in the distribution of geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?

@@

Our understanding of the patterns of cultural and genetic diversity among the early modern humans of the Upper Paleolithic remains incomplete. Previous works have performed various cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity. The studies have primarily relied upon personal ornamentation and parietal art, with few taking mobile objects into account.

@@

Our use of geometric signs in this study is supported by @Sauvet_2018, in which used quadrilateral geometric parietal signs are used to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain. The authors identified and classified the signs based on variations in form, then performed a factorial correspondence analysis and ascending hierarchical clustering upon the data in order to identify two clusters, which they argued represents distinct and separate groups of humans. In their study, @Sauvet_2018 made a case for the relevance of signs as identity markers, pointing to both the relevant qualities of signs, which are similar to those explained above, and the case studies throughout history where they served a similar purpose.

@@

abundance_graph <- ggplot(abundance_data) +
  aes(x = sign_total, y = reorder(site_name, sign_total)) +
  geom_col() +
  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) +
  labs(x = "Site name",
       y = "Sign count") +
  theme(axis.text.y = element_text(size = 4.25))

library(cowplot)
plot_grid(date_graph, abundance_graph, artifact_map,
          labels = "AUTO", align = "v",
          nrow = 3, scale = c(0.75, 0.75, 1.25))

@@

The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance. Assuming that this framework is true, we would hypothetically see patterns of cultural features of Upper Paleolithic early modern humans that are dispersed according to geography. Conversely, if geographic distance cannot account for the distribution of the sign types, we would have to assume that other factors are at play, namely, movement of groups or the creation of social networks.

@@

In this paper, we set out to ask if identifiable and relevant patterns existed in the data, and if so, what could they tell us about the patterns of diversity and movement among Upper Paleolithic humans?

@@

However, this preliminary research has been able to produce some conclusions to be reasonably confident in. Firstly, given that we know definitively that distribution and variation of sign types is not the result of geographic distance, we can conclude that the data is likely relevant to understanding patterns of cultural diversity. Secondly, we have found clusters of sites that correspond in sign type, possibly indicating mobility of specific cultural groups. These conclusions contribute to, and broaden our current understanding of cultural and biological diversity among Upper Paleolithic Homo sapiens in Europe.
